\section{Current Status and Roadmap}
\label{sec:vayu_roadmap}

\subsection{Current Status}

Vayu has reached a functional stage where the core flight control pipeline is fully implemented and operational. The system supports end-to-end execution from sensor acquisition to motor actuation, enabling closed-loop control suitable for real-time applications.

The current implementation includes:
\begin{itemize}
    \item Sensor integration using BMX160 (9-axis IMU) and BMP180 (barometer)
    \item State estimation using complementary and Mahony filters
    \item Cascaded control system with rate (1\,kHz) and attitude (250\,Hz) loops
    \item PWM-based motor actuation using hardware timers
    \item RC input via iBus protocol with DMA-based reception
    \item Structured telemetry system with multi-rate data streaming
    \item Persistent logging using VFS with SD card support
    \item Preemptive RTOS-based execution using VAIOS
\end{itemize}

The system is capable of stable real-time operation with modular task-based execution. Core subsystems have been tested in isolation and integrated environments, and the overall architecture has been validated through system-level execution.

\subsection{Roadmap}

The development of Vayu is ongoing, with focus on improving robustness, feature completeness, and system scalability. The roadmap is structured into short-term and long-term goals.

\textbf{Short-Term Goals:}
\begin{itemize}
    \item Hardware validation through flight testing
    \item Sensor calibration improvements and automated routines
    \item Tuning and optimization of control loops
    \item Enhanced failsafe mechanisms and safety features
    \item Improved telemetry visualization and ground interface tools
\end{itemize}

\textbf{Mid-Term Goals:}
\begin{itemize}
    \item Integration of additional sensors (e.g., GPS, magnetometer refinement)
    \item Advanced state estimation techniques
    \item Flight data recording (blackbox logging)
    \item Support for additional communication interfaces
\end{itemize}

\textbf{Long-Term Goals:}
\begin{itemize}
    \item Autonomous flight capabilities and navigation stack
    \item Multi-vehicle coordination and networking
    \item Expansion to multiple hardware platforms
    \item Integration with higher-level mission planning systems
\end{itemize}

\subsection{Development Direction}

The long-term vision of Vayu is to provide a fully indigenous, modular, and high-performance flight control stack that can be adapted across a wide range of applications. Development will continue to focus on maintaining a balance between performance, simplicity, and extensibility, ensuring that the system remains both practical and scalable.
